---
title: "Introduction to anchorCorrectR"
author:
  - María Rivas-Torrubia
  - Ana Brescia-Zapata
  - Guillermo Barturen
package: anchorCorrectR
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to anchorCorrectR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  dpi = 96
)
has_glmnet <- requireNamespace("glmnet", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

# Introduction

Technical batch effects are common in multi-batch RNA-seq studies. When the
same biological samples (**anchors**) are profiled in more than one batch,
those replicates provide a direct estimate of technical shift.

**anchorCorrectR** estimates batch effects from anchors and corrects
gene-expression matrices with:

- **Location shift** (`method = "shift"`): mean- or median-based offsets
- **Ridge regression** (`method = "ridge"`): regularized batch coefficients
  fitted on anchors only
- **Anchor-aware ComBat** (`method = "combat"`): location/scale with empirical
  Bayes shrinkage

This vignette follows the worked **SEQC / GSE47774** example from the package
README: platform is the batch, Sample (A/B/C/D) is biology, and cross-platform
C/D IDRs are anchors. Full scenario tables, ComBat-seq comparisons, and QC
examples live in the [GitHub README](https://github.com/bioinfoUGR/AnchorCorrectR).

```{r load}
library(anchorCorrectR)
```

# Load the SEQC example

Bundled counts and metadata are in `inst/extdata/`:

```{r load-data}
counts <- as.matrix(read.delim(
  system.file("extdata", "counts.tsv.gz", package = "anchorCorrectR"),
  row.names = 1, check.names = FALSE
))
meta <- read.delim(
  system.file("extdata", "metadata.tsv.gz", package = "anchorCorrectR"),
  check.names = FALSE
)
stopifnot(identical(colnames(counts), meta$ID))

dim(counts)
table(meta$Platform, meta$Sample)
```

| Column | Role |
|--------|------|
| `Platform` | Batch (`ILLUMINA` vs `ABI_SOLID`) |
| `Sample` | Biology (`A`, `B`, `C`, `D`) |
| `IDR` | Replicate ID (C/D on both platforms → anchors) |
| `ID` | GEO accession (= count matrix columns) |

# Design: 5 C/D anchors + balanced A/B

Keep five cross-platform C/D anchors and 30 A/B libraries with roughly even
platform balance (`scenario = "balanced"`). A/B libraries are **not** anchors,
so `assess_correction()` can still compute PCA-based metrics after excluding
replicate C/D samples.

```{r design}
set.seed(123)

cd <- meta[meta$Sample %in% c("C", "D"), ]
cd_anchor_ids <- names(which(tapply(cd$Platform, cd$IDR, function(z) {
  all(c("ILLUMINA", "ABI_SOLID") %in% z)
})))
keep_idr <- sample(cd_anchor_ids, 5)
meta_cd <- cd[cd$IDR %in% keep_idr, ]
keep_idr

pick_ab <- function(meta_ab, sample_label, n_ilm, n_solid) {
  ilm <- meta_ab$ID[meta_ab$Sample == sample_label &
                      meta_ab$Platform == "ILLUMINA"]
  sol <- meta_ab$ID[meta_ab$Sample == sample_label &
                      meta_ab$Platform == "ABI_SOLID"]
  c(sample(ilm, n_ilm), sample(sol, n_solid))
}

ab <- meta[meta$Sample %in% c("A", "B"), ]
ids_ab <- c(pick_ab(ab, "A", 8, 7), pick_ab(ab, "B", 8, 7))

meta_use <- rbind(meta_cd, ab[ab$ID %in% ids_ab, ])
meta_use <- meta_use[match(intersect(meta_use$ID, colnames(counts)),
                           meta_use$ID), ]
counts_use <- counts[, meta_use$ID, drop = FALSE]

batch     <- factor(meta_use$Platform)
biology   <- factor(meta_use$Sample)
sample_id <- factor(meta_use$IDR)

table(batch, biology)
```

Other README scenarios (`mild`, `strong`, `confounded`) only change the A/B
platform counts inside `pick_ab()`.

# Correct with shift and anchor-aware ComBat

```{r correct}
out_shift <- anchor_correct(
  counts_use, batch, sample_id,
  method = "shift", center = "mean",
  input_type = "counts", ref_batch = "ILLUMINA",
  verbose = FALSE
)

out_combat <- anchor_correct(
  counts_use, batch, sample_id,
  method = "combat",
  input_type = "counts", ref_batch = "ILLUMINA",
  verbose = FALSE
)

dim(out_shift)
```

Use `center = "median"` when anchors may be noisy. Optional `ref_batch` leaves
one platform uncorrected (here Illumina).

```{r ridge, eval=has_glmnet}
out_ridge <- anchor_correct(
  counts_use, batch, sample_id,
  method = "ridge",
  input_type = "counts", ref_batch = "ILLUMINA",
  verbose = FALSE
)
```

```{r ridge-skip, eval=!has_glmnet, results='asis'}
cat("_Skipping ridge: install Suggested package `glmnet` to run this chunk._\n")
```

# Assess and compare methods

`assess_correction()` reports PVCA batch/biology variance, anchor RMSE, kNN
Jaccard, HVG overlap, and clustering concordance. With `sample_id` set, C/D
anchors are excluded from PCA-based metrics (A/B remain) but still used for
RMSE.

```{r assess}
a_shift <- assess_correction(
  counts_use, out_shift, batch, biology, sample_id,
  input_type = "counts", verbose = FALSE
)
a_combat <- assess_correction(
  counts_use, out_combat, batch, biology, sample_id,
  input_type = "counts", verbose = FALSE
)
```

```{r merge-shift-combat, eval=!has_glmnet}
merged <- merge_assessments(
  a_shift, a_combat,
  run_names = c("shift", "combat")
)
merged$consolidated_table
```

```{r merge-with-ridge, eval=has_glmnet}
a_ridge <- assess_correction(
  counts_use, out_ridge, batch, biology, sample_id,
  input_type = "counts", verbose = FALSE
)
merged <- merge_assessments(
  a_shift, a_ridge, a_combat,
  run_names = c("shift", "ridge", "combat")
)
merged$consolidated_table
```

On this balanced SEQC subset, all package methods reduce platform variance and
roughly halve C/D anchor RMSE while raising biology variance. For imbalance /
confounding designs and an `sva::ComBat_seq` baseline, see the README.

# MDS before / after

```{r mds, eval=has_ggplot2, fig.cap="MDS of SEQC samples before and after mean-shift correction (balanced design). Color = platform; shape = Sample."}
plot_mds_before_after(
  counts_use, out_shift, batch, biology,
  input_type = "counts",
  title = "MDS before vs after — shift · balanced"
)
```

```{r mds-skip, eval=!has_ggplot2, results='asis'}
cat("_Skipping MDS plot: install Suggested package `ggplot2`._\n")
```

# Notes

- Input must be a numeric **genes × samples** matrix (`input_type = "counts"`
  or `"log"`).
- For `SummarizedExperiment` objects, extract `assay()` / `colData()` columns
  and pass them to `anchor_correct()`.
- Extended QC helpers (`detect_outliers_pca_mds()`, `detect_sex_mismatch()`,
  `vcf_compare_fast()`, `generate_bootstrap_fastq()`) are documented in the
  README with PRECISESADS cohort examples.

# Session info

```{r sessionInfo}
sessionInfo()
```
