---
title: "Introduction to anchorCorrectR"
author:
  - María Rivas-Torrubia
  - Ana Brescia-Zapata
  - Guillermo Barturen
package: anchorCorrectR
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to anchorCorrectR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Introduction

Technical batch effects are common in multi-batch RNA-seq and microarray
studies. When the same biological samples (anchors) are profiled in more than
one batch, those replicates provide a direct estimate of technical shift.

**anchorCorrectR** uses such anchors to correct gene-expression matrices with:

- **Location shift** (`method = "shift"`): mean- or median-based offsets
- **Ridge regression** (`method = "ridge"`): regularized batch coefficients
- **Anchor-aware ComBat** (`method = "combat"`): location/scale with empirical Bayes shrinkage

Related Bioconductor / CRAN tools include `sva::ComBat`,
`limma::removeBatchEffect`, and `batchelor`. Those methods typically estimate
batch effects from all samples in a batch. anchorCorrectR instead estimates
effects from cross-batch replicates, which is useful when biology and batch are
partially confounded.

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
# After Bioconductor acceptance:
# BiocManager::install("anchorCorrectR")

# From GitHub during development:
# BiocManager::install("gbarturen/anchorCorrectR")
```

```{r load}
library(anchorCorrectR)
```

# Simulated multi-batch experiment

```{r simulate}
set.seed(123)
n_genes <- 200
n_per_batch <- 10
n_batches <- 3

sample_id <- factor(rep(seq_len(n_per_batch), times = n_batches))
batch <- factor(rep(paste0("Batch", seq_len(n_batches)), each = n_per_batch))
biology <- factor(rep(rep(c("GroupA", "GroupB"), each = n_per_batch / 2),
                      times = n_batches))

counts <- matrix(
  rpois(n_genes * length(sample_id), lambda = 80),
  nrow = n_genes,
  dimnames = list(
    paste0("Gene", seq_len(n_genes)),
    paste0("S", seq_along(sample_id))
  )
)

# Artificial multiplicative batch effects
batch_mult <- c(Batch1 = 1.0, Batch2 = 1.7, Batch3 = 0.65)
counts <- sweep(counts, 2, batch_mult[as.character(batch)], `*`)
```

# Location-shift correction

```{r shift}
corrected_mean <- anchor_correct(
  counts, batch, sample_id,
  method = "shift",
  center = "mean",
  input_type = "counts",
  verbose = FALSE
)

corrected_median <- anchor_correct(
  counts, batch, sample_id,
  method = "shift",
  center = "median",
  input_type = "counts",
  verbose = FALSE
)

dim(corrected_mean)
```

Use `center = "median"` when a few anchors may be unreliable. Optionally keep
one batch unchanged with `ref_batch`.

```{r ref-batch}
corrected_ref <- anchor_correct(
  counts, batch, sample_id,
  method = "shift",
  center = "mean",
  ref_batch = "Batch1",
  input_type = "counts",
  verbose = FALSE
)
```

# Anchor-aware ComBat

```{r combat}
corrected_combat <- anchor_correct(
  counts, batch, sample_id,
  method = "combat",
  input_type = "counts",
  verbose = FALSE
)
```

# Assessing correction

```{r assess}
assessment <- assess_correction(
  x_before = counts,
  x_after = corrected_mean,
  batch = batch,
  biology = biology,
  sample_id = sample_id,
  input_type = "counts",
  verbose = FALSE
)
print(assessment)
```

MDS before versus after correction:

```{r mds, fig.cap="MDS of samples before and after shift correction."}
plot_mds_before_after(
  x_before = counts,
  x_after = corrected_mean,
  batch = batch,
  biology = biology,
  input_type = "counts"
)
```

# Working with matrix inputs

anchorCorrectR expects a numeric **genes × samples** matrix. For
`SummarizedExperiment` / `SingleCellExperiment` objects, extract the assay and
metadata columns, then correct:

```{r se-pattern, eval=FALSE}
# library(SummarizedExperiment)
# se <- ...
# corrected <- anchor_correct(
#   assay(se, "counts"),
#   batch = colData(se)$batch,
#   sample_id = colData(se)$sample_id,
#   method = "shift",
#   center = "mean",
#   input_type = "counts"
# )
# assay(se, "corrected") <- corrected
```

# Session info

```{r sessionInfo}
sessionInfo()
```
